\section{Conclusion}
\label{s:conclusion}

Until recently, it was widely assumed that the fastest switching chips would be
fixed-function; a programmable device could not have the same performance.
Recent research into programmable parsers~\cite{glen_parsing}, fast
programmable switch pipelines~\cite{rmt}, and languages to program
them~\cite{p4, pof}, coupled with recent multi-Tbit/s programmable commercial
chips~\cite{tofino, xpliant} suggests that change might be afoot.

But so far,
it has been considered off-limits to program the packet scheduler---in part
because the desired algorithms are so varied, and because the scheduler sits at
the heart of the shared packet buffer where timing requirements are tightest.
It has been widely assumed too hard to find a useful abstraction that can also
be implemented in fast hardware.

PIFOs appear to be a very promising abstraction: they include a variety of
existing algorithms, and allow us to express new ones. Further, they can be
implemented at line rate with modest chip area overhead.

We believe the most
exciting consequence will be the creation of many new schedulers, invented by
network operators, iterated and refined, then deployed for their own needs. No
longer will research experiments be limited to simulation and progress
constrained by a vendor's choice of scheduling algorithms. Those needing a new
algorithm could create their own, or even download one from an open-source
repository or a reproducible SIGCOMM paper.

To get there, we will need real switching chips with programmable PIFO
schedulers. The good news is that we see no reason why future switching chips
can not include a programmable PIFO scheduler.
